The Talking Heads Experiment

### 2015-04-17 chap12/chap12.tex
orphans, widows, line breaks

Today, Darwinian evolutionary dynamics \is{Darwinian evolutionary dynamics}
is recognised as a general principle that can and has been applied to 
many different types of systems, in chemistry, neurobiology or 
economics.\footnote{See \cite{Arthur:1996}, \cite{Edelman:1987} and \cite{Luisi:2001}.}
It has also been used successfully to come up 
with new artificial systems, for example to evolve controllers for robots using 
genetic algorithms \citep{Nolfi:2000}. It is therefore not so surprising that 
selectionism can also be mapped to the cultural and more specifically to the linguistic domain in the following way: 
\begin{enumerate}
\item Speakers generate variants, either accidentally through errors in their speech production, or deliberately, 
for example, when they try to express a novel idea by inventing new forms or 
appropriating existing forms for new purposes, for example, by coercing a word or grammatical contruction into a new usage. 
\item Which variants will survive in the communal language depends on selectionist criteria. I will 
discuss them shortly in more detail. They include for example whether the speaker is understood by the hearer, or whether 
it reduces the cognitive complexity of parsing and producing and hence allows larger sentences to be understood. 
\item The self-enforcing causal loop happens when variants created by one speaker start to propagate in the population 
and begin to replace their competitors. For example, a particular word to express a given meaning gets out of fashion 
and is taken over by another one. Some of these developments are simply fashion trends, but the more profound ones, 
in particular the ones that involve grammar, are related to critical selectionist criteria. 
\end{enumerate}
This Darwinian dynamic is already active in the operation of language strategies, because a strategy regulates the 
creation and selection of variants, determines how feedback has to be handled, and implies a self-enforcing causal loop through 
lateral inhibition so that variants satisfying the selection criteria become 
dominant in the population.

@@

By analogy with biology, we can apply the concept of a niche to the evolutionary study of
language systems and language strategies. To investigate linguistic niches in a systematic way, we need to identify first 
what factors are relevant. Some of these factors 
have to do with external conditions, for example, if the language games require expressing fine-grained hue distinctions or 
temporal and aspectual distinctions. Others relate to internal conditions, for example, how big is the search space during parsing, 
how much does a model of the situation need to be consulted. Next we can 
examine the performance of different systems and strategies for these factors, and map performance as 
points or regions in the niche space. Then we can study whether these points move towards greater or lesser optimality over time, 
how different systems or strategies compete when occupying the same niche, or how language systems and language 
strategies adapt to changes in the ecological conditions and hence the dimensions or critical values
of the niche space. The factors relevant for a linguistic niche include the selectionist criteria discussed earlier, 
and they can be analysed by identifying a fitness landscape, where each dimension is a selection component. 
I discuss first examples of this kind of analysis for language systems and then for language strategies. \\

@@

Van Trijp initialised three populations of agents: with the Old High German system, 
the Middle High German system, or the New High German System. He then defined the following four dimensions for the 
fitness landscape: 
\begin{enumerate}
\item {\itshape Search effort} which is the amount of search that is needed, scaled between minimum (when there is one marker for every of 
the 18 possible configurations and so no search is needed) and maximum (when there are no markers at all).
\item {\itshape Disambiguation power} which is based on the number of possible interpretations left after application of the grammar. 
The disambiguation power is also scaled with maximum power for 12 markers and minimum for 0 markers (not counting the genitive). 
\item {\itshape Articulatory effort} which is calculated in terms of the number of movements that articulators such as the lips and 
tongue have to make when pronouncing the different phonemes that make up each article. 
\item {\itshape Acoustic precision} which is based on the phonological distance between words.
\end{enumerate}
Criteria (1), (2) and (4) pertain to the hearer and (1) and (3) to the speaker. They are calculated for each utterance
in a corpus and the weighted average gives a value for the performance of the language system for each of these dimensions. The global 
performance can be assessed by taking a (weighted) average of each of the values for all dimensions. 
As expected, Old High German scores the highest for disambiguation power and acoustic precision and the lowest for articulatory effort 
and processing effort, thus being an advantage for the hearer but not for the speaker (see \figref{fig:german-system}).

@@

This example shows how it is possible to analyse the fitness landscape for stages of a language system at different points in 
time. The same analysis can also be done for different regional variants that are currently competing to become dominant 
in the standard language. An example of the latter is the currently ongoing evolution
from the etymological Spanish pronoun system to the referential pronoun systems, 
which has three different possible regional variants: `le\'ismo', `la\'ismo' and `lo\'ismo'. 
Van Trijp showed how each of these variants could equally arise with the same referential strategy and how it is 
a matter of historical contingency which one becomes dominant. Indeed we see that different regions of Spain
have chosen different variants \citep{VanTrijp:2010}. 
\clearpage
Here is another example of a fitness landscape from the domain of space, developed by 
Michael Spranger.\footnote{These results are documented at length in: \cite{Spranger:2014}.}
It focuses on understanding the role of grammar for optimising spatial language. \is{Spatial Language Game}
The experiment involves \textsc{qrio} humanoids
(see \figref{fig:spatial}) and uses a 
sophisticated vision system developed by Spranger, together with IRL and Fluid 
Construction Grammar (in the previous chapter). A variety of spatial language strategies 
were operationalised, in particular:

@@

Spranger analysed the fitness landscape for grammatical strategies. He compared 
one population of agents that was initialised with a pidgin German, i.e. only a vocabulary for objects and spatial relations 
but no grammar, and another population initialised with a full locative grammar of German. He investigated three external 
factors: 
\begin{enumerate}
\item Scenes with many objects distributed around a landmark.
\item Scenes with several objects but no landmark that can be used as a spatial vantage point. 
\item Scenes with an allocentric and without an allocentric landmark but more complex with respect to number 
of objects and distribution of objects compared to (1). 
\end{enumerate}

@@

Let us focus on a family of strategies that all use a prototype-based approach, i.e. they perform categorisation using 
a standard nearest neighbour classification algorithm operating over an inventory of prototypical points, 
as also used in the Proper Naming Game and the Basic Colour Naming game discussed in the previous chapter. 
Three variant strategies are imaginable (and can be found in human languages): 
\begin{enumerate}
\item The {\itshape brightness-only strategy} which uses only the L* dimension (dark, grey, bright, etc.).
\item The {\itshape hue-only strategy} which uses only the hue dimension (blue, yellow, green, etc.). 
\item And the {\itshape full-color strategy} which uses all dimensions. 
\end{enumerate}
In order to understand the niche of each strategy, Bleys has set up experiments where the context is controlled 
with respect to the importance of hue. 
For example, agents can be presented with contexts with samples where the hue is the same but brightness is different on the one 
hand versus contexts with samples where the brightness is the same but hue is different. In between we get a mixed 
importance of hue.

@@

Recall that three internal factors were used 
for the fitness landscape in this domain: communicative success, semantic ambiguity and interpretation sharing. Spranger started from 
a population without any grammar but just a lexicon for expressing spatial relations and properties of objects. 
The agents were then initialised with a strategy that would build a grammar consisting of hierarchical
constructions which group all the components of a spatial relation, put them in a particular sequential order,
and add a marker to them. For example, a construction could arise that groups a spatial relation (like `front'), 
a determiner acting as selector (like `the'), and a noun introducing an object class (e.g. `box'), puts them 
in the order: spatial relation - determiner - noun, and adds a marker to them (e.g. `-bo').

@@

A series of language games were played using contexts (such as the one shown in \figref{fig:gramdev3}, left)
that contained a wide range of objects and also a landmark stimulating the need for perspective reversal. As shown in 
the niche analysis earlier (see Figure \ref{fig:ambiguity} and \ref{fig:why-grammar-german}) a language without 
a grammatical expression supporting how spatial relations need to be analysed will reach a 
communicative success rate of only 70 \%. So the question is whether the agents will increase this success rate when 
they exercise this grammatical strategy. The same question can be posed for the other two niche factors (disambiguation 
and semantic sharing). Experimental results are shown in \figref{fig:gramdev3}. They confirm that as 
the language system is getting built up, key factors (communicative success, semantic sharing, and disambiguation) 
indeed get optimised, demonstrating that the strategy ``does its job''.

\section{Selection and alignment of language strategies}

@@

\begin{figure}[htb!]
\centerline{
\includegraphics[width=0.45\textwidth]{chap12/figs/full-cs-graph.pdf}
\includegraphics[height=0.3\textwidth]{chap12/figs/full-1000.pdf}$~~~$
\includegraphics[height=0.3\textwidth]{chap12/figs/full-3000.pdf}
}
\caption{{The {\itshape Basic Colour Strategy} allows a population of agents (in this case 10) to self-organise a colour lexicon from scratch. The graph shows that high communicative success is reached with a lexicon of about 15 colour words (left). The evolution of a typical lexicon in a smaller population (5 agents), is shown after 400 (middle) and 1200 (right) games per agent. Each row represents the lexicon of one agent.}
\label{fig:full-cs-dynamics}
} 
\end{figure}

@@

\begin{figure}[htb!]
\centerline{
\includegraphics[width=0.4\textwidth]{chap12/figs/brightness-graph.pdf}
\includegraphics[height=0.3\textwidth]{chap12/figs/bw-1000.pdf} $~~~$
\includegraphics[height=0.3\textwidth]{chap12/figs/bw-4000.pdf}}
\label{fig:brightness-dynamics}
\caption{{The {\itshape Brightness Colour Strategy} also allows a population of agents (in this case 10)
to evolve an adequate colour lexicon (left). The evolution of a typical lexicon in a smaller population (5 agents), is shown 
after 400 (middle) and 1600 (right) games per agent. Each row represents
the lexicon of one agent.}}\todo[inline]{barely readable at 100\%. Do we need to change s.th. here?}
\end{figure}

@@

\begin{figure}[hp]
\centerline{\includegraphics[width=\textwidth]{chap12/figs/strategies-one-winner.pdf}}
\caption{{Communicative success stays constant. The Brightness Colour Strategy becomes 
dominant very early due to random fluctuations. It stays dominant but barely. 
The strategy coherence (i.e. whether agents agree on the 
default strategy) increases gradually. The Full Colour Strategy remains still in use for cases where
it is needed. \label{fig:strategies-one}}}
\end{figure}

@@

\begin{figure}[hp]
\centerline{\includegraphics[width=\textwidth]{chap12/figs/strategies-competition.pdf}}
\caption{{Initially the brightness strategy is dominant, but then the full-colour 
strategy gradually takes over and becomes dominant around 400 games/agents. Strategy coherence within the population
progressively increases.}
\label{fig:strategies-dynamics}}
\end{figure}

@@

The previous section showed how there can be competition or cooperation between strategies, fought out through the 
performance of the language systems against the niche factors in which \is{strategy generation}
the language strategies operate. But a Darwinian evolutionary framework requires that there is not only 
selection and a self-enforcing causal feedback loop 
but also that new elements get generated, in this case new strategies, which can then be subjected to 
a selectionist dynamics. How can we explain that new 
strategies arise? This is the most challenging difficult question of evolutionary linguistics research and very much an 
open question. We basically have to move up to a meta-level where we have strategies for building 
strategies, further called meta-strategies. \is{meta-strategy} It is unclear whether they have to be innately given or 
whether they are themselves also the result of developmental processes and a cultural dynamics. 
There are two aspects to a strategy: a conceptual side and a linguistic side. For the conceptual 
side, solid results have been achieved using IRL, which was from the beginning developed to allow the generation of 
new conceptualisation strategies. So far a single meta-strategy has proven to be quite successful in this case. 
The linguistic side however is still to a large extent in a stage of fundamental research, and here there appear to 
be several different meta-strategies, used by different language communities. Each of these two aspects is now 
discussed.\\

@@

\begin{figure}
\centerline{\includegraphics[width=1.0\linewidth]{chap11/figs/near-the-box2.pdf}}
\caption{Example of a conceptualisation for the phrase `near to the box'. {\itshape Top:} 
Leftmost: The situation as perceived by the robot. Middle: situation after a geometric transform where the box is 
made the central perspective. Right: Area around the box which reflects (in terms of darkness of the color) how 
near objects are to the box. {\itshape Bottom:} IRL constraint network for this conceptualisation.
\label{fig:spatial-strategies}}
\end{figure}

We will use the domain of spatial language again as a source of examples, based on experiments carried out by 
Michael Spranger. It uses the \textsc{qrio} humanoid robots 
and scenes like the one shown in \figref{fig:spatial}. An example of a constraint network from this domain 
is shown in \figref{fig:spatial-strategies}. It is the network for the phrase `near the box'. `Near' is a proximal 
relation which categorises objects in terms of how near they are to a landmark (in this case the box). In a list form the 
network looks like this: 
\begin{verbatim}
1. (get-context ?ctx-1) 
2. (construct-region-proximal 
    ?region ?ctx-1 ?landmark ?cat) 
3. (bind proximal-category ?cat NEAR) 
4. (apply-selection ?landmark ?landmarks ?selector-2)
5. (apply-class ?landmarks ?ctx ?landmark-class) 
6. (bind selector ?selector-2 UNIQUE) 
7. (get-context ?ctx) 
8. (bind object-class ?landmark-class BOX)
\end{verbatim}
The order of these operations does not matter as values propagate and constraints become active until no further progress
can be made. Operations in 4, 5, 6, 7 and 8 provide the referent of `the box'. The following contributions 
can be made by each of the cognitive operations: The predicate BOX gets bound to ?landmark-class in operation 8. Object-class is 
the type of ?landmark-class. The context, i.e. all objects in the present situation, gets bound to ?ctx in operation 7. 
The selector UNIQUE gets bound to ?selector-2 in operation 6. 
Then the operation apply-class is applied to all the objects in ?ctx (the context) to find the possible landmarks which 
get bound to ?landmarks using the ?landmark-class (BOX). This set is assumed to be a singleton and so 
in operation 4 the unique element in the set of ?landmarks can be selected 
and bound to ?landmark.

The remaining three operations apply the proximal relation. In operation 3 ?cat is bound to the proximal-category NEAR. 
In operation 1, ?ctx-1 gets bound to 
the context. And finally in operation 2, the region around the landmark, which satisfies the category ?cat (NEAR), can 
be computed and bound to ?region by ``construct-region-proximal''. The construction-region-proximal constraint has to perform first 
a geometric transformation on the scene, constructing a world model from the viewpoint of the landmark, i.e. the box.

@@

\figref{fig:all} shows an example experiment in which all the aspects discussed in this section are brought together. 
A population of 10 agents is developing new conceptualisation strategies for a niche that encourages preference 
for a particular strategy, namely a projective strategy with possible perspective reversal, as in `the box in front 
of the wall'. We see that communicative success steadily rises and that the number of 
spatial categories increases. The bottom graph zooms in on the conceptualisation strategies. We see that agents reach 
a consensus about a dominant strategy after a period in which new strategies 
are invented and tried out.\\

\subsection{Meta-strategies for generating new lexico-gram\-matical strategies}
% {\bfshape Meta-strategies for generating new lexico-grammatical strategies}\\

Are there general meta-strategies that can be used to generate the kind of lexico-grammatical strategies 
that can in turn generate all the language systems we observe 
in today's languages? This is a very profound question, related to the ongoing debate in linguistics on the existence of 
a universal language acquisition device. A lot of ink has been spilled but we need to do profound computational modeling 
of meta-strategies and thorough experiments to find out. 
\clearpage
Generally speaking, a lexico-grammatical strategy has two components: {\itshape diagnostics} which monitor how 
language processing is going and {\itshape repairs} which add new 
constructions to the agent's inventory.

@@

It is not really possible to capture the enormous creativity of human word creation in computational simulations 
because that would involve components for common sense inference and access to massive cultural knowledge. 
However we can do experiments to understand how new words may arise by analogy with existing words 
and how the meaning of words can develop further, for example by abandoning some of the aspects of 
the original meaning or gaining new meanings. Such experiments have 
been carried out in particular by \cite{Wellens:2008}. \is{flexible word meaning}

@@

First we need much more work along the paths of which we have seen many examples: 
\begin{enumerate}
\item The study of more language systems, based on firm empirical data. For example, the study of the verbal system of 
Italian or the determiner system of Dutch. 
\item A study of the language strategies that underly these systems. A language strategy should define how a particular 
type of language system operates, how it is learned, and how it is expanded and aligned. 
\item A study of the niche occupied by the language systems governed by a particular language strategy. 
\item Evolutionary experiments to show how new strategies may arise. 
\end{enumerate} 
Then there are also several deep open questions which have not been addressed extensively so far. Here are a few examples: 
\begin{enumerate}
\item What are implications of the models presented here for the neurobiology of language? Right now it remains largely 
a mystery how the human brain is processing language and current neural network models are entirely insufficient from 
a representational and computational point of view to handle the structures and processes that we have been using in 
the various experiments reported in this book. 
\item How to orchestrate long-term evolution, both within the individual and in a population? Some suggestions have been 
made based on the autotelic principle which allows agents to control the communicative challenges in 
relation to their level of skill \citep{Steels:2007}. 
But these suggestions need to be tested
more seriously in large-scale long term experiments. 
\item What is the interaction between cultural evolution and genetic evolution? I have argued in favor of culture-driven genetic
evolution but what kind of genetic change is necessary to support the processes that we have been found to 
be operating at the cultural level? Clearly, flexible recruitment and huge amounts of memory are key demands on the 
human brain, but there are undoubtly others. 
\item What is the interaction between social evolution, cultural evolution and biological evolution? The language 
games which we have studied all required honest communication. Agents do their best to play the game as well as they 
can and agents are fully cooperative. This is not the default condition in the natural world and it is not so easy 
to explain how human sociality arose, and how human language may have fostered it \citep{Dor:2014}. 
\end{enumerate}
These questions can only be properly addressed by bringing up and working out hypotheses and
doing detailed computational analysis and computational 
experiments. The field of evolutionary linguistics is still in its infancy and so far too few researchers
are contributing through the kind of computational simulations that were discussed in this book. This is not surprising given 
the technical complexity and the breadth of knowledge needed to work in this area. It takes Ph.D students on average 5
years of training and experimentation to carry a research thread to 
fruition, assuming that the student starts with considerable background in computational linguistics or Artificial Intelligence.
Let us hope that in the future the resources will be available so that younger generations can pursue this exciting path. 
Many discoveries are waiting to be made!

>>>>

Today, Darwinian evolutionary dynamics \is{Darwinian evolutionary dynamics}
is recognised as a general principle that can and has been applied to 
many different types of systems, in chemistry, neurobiology or 
economics.\footnote{See \cite{Arthur:1996}, \cite{Edelman:1987} and \cite{Luisi:2001}.}
It has also been used successfully to come up 
with new artificial systems, for example to evolve controllers for robots using 
genetic algorithms \citep{Nolfi:2000}. It is therefore not so surprising that 
selectionism can also be mapped to the cultural and more specifically to the linguistic domain in the following way: 
\begin{enumerate}
\item Speakers generate variants, either accidentally through errors in their speech production, or deliberately, 
for example, when they try to express a novel idea by inventing new forms or 
appropriating existing forms for new purposes, for example, by coercing a word or grammatical contruction into a new usage. 
\item Which variants will survive in the communal language depends on selectionist criteria. I will 
discuss them shortly in more detail. They include for example whether the speaker is understood by the hearer, or whether 
it reduces the cognitive complexity of parsing and producing and hence allows larger sentences to be understood. 
\item The self-enforcing causal loop happens when variants created by one speaker start to propagate in the population 
and begin to replace their competitors. For example, a particular word to express a given meaning gets out of fashion 
and is taken over by another one. Some of these developments are simply fashion trends, but the more profound ones, 
in particular the ones that involve grammar, are related to critical selectionist criteria. 
\end{enumerate}
This Darwinian dynamic is already active in the operation of language strategies, since \todo{replaced `because' with `since' here to fit the line} a strategy regulates the 
creation and selection of variants, determines how feedback has to be handled, and implies a self-enforcing causal loop through 
lateral inhibition so that variants satisfying the selection criteria become 
dominant in the population.

@@

By analogy with biology, we can apply the concept of a niche to the evolutionary study of
language systems and language strategies. To investigate linguistic niches in a systematic way, we need to identify first 
what factors are relevant. Some of these factors 
have to do with external conditions, for example, if the language games require expressing fine-grained hue distinctions or 
temporal and aspectual distinctions. Others relate to internal conditions, for example, how big is the search space during parsing, 
how much does a model of the situation need to be consulted. Next we can 
examine the performance of different systems and strategies for these factors, and map performance as 
points or regions in the niche space. Then we can study whether these points move towards greater or lesser optimality over time, 
how different systems or strategies compete when occupying the same niche, or how language systems and language 
strategies adapt to changes in the ecological conditions and hence the dimensions or critical values
of the niche space. The factors relevant for a linguistic niche include the selectionist criteria discussed earlier, 
and they can be analysed by identifying a fitness landscape, where each dimension is a selection component. 
I discuss first examples of this kind of analysis for language systems and then for language strategies.

@@

Van Trijp initialised three populations of agents: with the Old High German system, 
the Middle High German system, or the New High German System. He then defined the following four dimensions for the 
fitness landscape: \enlargethispage{1\baselineskip}
\begin{enumerate}
\item {\itshape Search effort} which is the amount of search that is needed, scaled between minimum (when there is one marker for every of 
the 18 possible configurations and so no search is needed) and maximum (when there are no markers at all).
\item {\itshape Disambiguation power} which is based on the number of possible interpretations left after application of the grammar. 
The disambiguation power is also scaled with maximum power for 12 markers and minimum for 0 markers (not counting the genitive). 
\item {\itshape Articulatory effort} which is calculated in terms of the number of movements that articulators such as the lips and 
tongue have to make when pronouncing the different phonemes that make up each article. 
\item {\itshape Acoustic precision} which is based on the phonological distance between words.
\end{enumerate}
Criteria (1), (2) and (4) pertain to the hearer and (1) and (3) to the speaker. They are calculated for each utterance
in a corpus and the weighted average gives a value for the performance of the language system for each of these dimensions. The global 
performance can be assessed by taking a (weighted) average of each of the values for all dimensions. 
As expected, Old High German scores the highest for disambiguation power and acoustic precision and the lowest for articulatory effort 
and processing effort, thus being an advantage for the hearer but not for the speaker (see \figref{fig:german-system}).

@@

This example shows how it is possible to analyse the fitness landscape for stages of a language system at different points in 
time. The same analysis can also be done for different regional variants that are currently competing to become dominant 
in the standard language. An example of the latter is the currently ongoing evolution
from the etymological Spanish pronoun system to the referential pronoun systems, 
which has three different possible regional variants: `le\'ismo', `la\'ismo' and `lo\'ismo'. 
Van Trijp showed how each of these variants could equally arise with the same referential strategy and how it is 
a matter of historical contingency which one becomes dominant. Indeed we see that different regions of Spain
have chosen different variants \citep{VanTrijp:2010}.

Here is another example of a fitness landscape from the domain of space, developed by 
Michael Spranger.\footnote{These results are documented at length in: \cite{Spranger:2014}.}
It focuses on understanding the role of grammar for optimising spatial language. \is{Spatial Language Game}
The experiment involves \textsc{qrio} humanoids
(see \figref{fig:spatial}) and uses a 
sophisticated vision system developed by Spranger, together with IRL and Fluid 
Construction Grammar (in the previous chapter). A variety of spatial language strategies 
were operationalised, in particular:

@@

Spranger analysed the fitness landscape for grammatical strategies. He compared 
one population of agents that was initialised with a pidgin German, i.e. only a vocabulary for objects and spatial relations 
but no grammar, and another population initialised with a full locative grammar of German. He investigated three external 
factors: \clearpage
\begin{enumerate}
\item Scenes with many objects distributed around a landmark.
\item Scenes with several objects but no landmark that can be used as a spatial vantage point. 
\item Scenes with an allocentric and without an allocentric landmark but more complex with respect to number 
of objects and distribution of objects compared to (1). 
\end{enumerate}

@@

Let us focus on a family of strategies that all use a prototype-based approach, i.e. they perform categorisation using 
a standard nearest neighbour classification algorithm operating over an inventory of prototypical points, 
as also used in the Proper Naming Game and the Basic Colour Naming game discussed in the previous chapter. 
Three variant strategies are imaginable (and can be found in human languages): \enlargethispage{2\baselineskip}
\begin{enumerate}
\item The {\itshape brightness-only strategy} which uses only the L* dimension (dark, grey, bright, etc.).
\item The {\itshape hue-only strategy} which uses only the hue dimension (blue, yellow, green, etc.). 
\item And the {\itshape full-color strategy} which uses all dimensions. 
\end{enumerate}
In order to understand the niche of each strategy, Bleys has set up experiments where the context is controlled 
with respect to the importance of hue. 
For example, agents can be presented with contexts with samples where the hue is the same but brightness is different on the one 
hand versus contexts with samples where the brightness is the same but hue is different. In between we get a mixed 
importance of hue.

@@

Recall that three internal factors were used \todo{replaced `started' with `began' here to fit the line}
for the fitness landscape in this domain: communicative success, semantic ambiguity and interpretation sharing. Spranger began with 
a population without any grammar but just a lexicon for expressing spatial relations and properties of objects. 
The agents were then initialised with a strategy that would build a grammar consisting of hierarchical
constructions which group all the components of a spatial relation, put them in a particular sequential order,
and add a marker to them. For example, a construction could arise that groups a spatial relation (like `front'), 
a determiner acting as selector (like `the'), and a noun introducing an object class (e.g. `box'), puts them 
in the order: spatial relation - determiner - noun, and adds a marker to them (e.g. `-bo').

@@

A series of language games were played using contexts (such as the one shown in \figref{fig:gramdev3}, left)
that contained a wide range of objects and also a landmark stimulating the need for perspective reversal. As shown in 
the niche analysis earlier (see Figure \ref{fig:ambiguity} and \ref{fig:why-grammar-german}) a language without 
a grammatical expression supporting how spatial relations need to be analysed will reach a 
communicative success rate of only 70 \%. So the question is whether the agents will increase this success rate when 
they exercise this grammatical strategy. The same question can be posed for the other two niche factors (disambiguation 
and semantic sharing). Experimental results are shown in \figref{fig:gramdev3}. They confirm that as 
the language system is getting built up, key factors (communicative success, semantic sharing, and disambiguation) 
indeed get optimised, demonstrating that the strategy ``does its job''. 
\clearpage
\section{Selection and alignment of language strategies}

@@

\begin{figure}[p]
\centerline{
\includegraphics[width=\textwidth]{chap12/figs/full-cs-graph.pdf}
}\\
\centerline{
\includegraphics[height=0.5\textwidth]{chap12/figs/full-1000.pdf}$~~~$
\includegraphics[height=0.5\textwidth]{chap12/figs/full-3000.pdf}
}
\caption{{The {\itshape Basic Colour Strategy} allows a population of agents (in this case 10) to self-organise a colour lexicon from scratch. The graph shows that high communicative success is reached with a lexicon of about 15 colour words (top). The evolution of a typical lexicon in a smaller population (5 agents), is shown after 400 (bottom left) and 1200 (bottom right) games per agent. Each row represents the lexicon of one agent.}
\label{fig:full-cs-dynamics}
} 
\end{figure}

@@

\begin{figure}[p]
\centerline{
\includegraphics[width=\textwidth]{chap12/figs/brightness-graph.pdf}
}\\
\centerline{
\includegraphics[height=0.5\textwidth]{chap12/figs/bw-1000.pdf} $~~~$
\includegraphics[height=0.5\textwidth]{chap12/figs/bw-4000.pdf}}
\label{fig:brightness-dynamics}
\caption{{The {\itshape Brightness Colour Strategy} also allows a population of agents (in this case 10)
to evolve an adequate colour lexicon (top). The evolution of a typical lexicon in a smaller population (5 agents), is shown 
after 400 (bottom left) and 1600 (bottom right) games per agent. Each row represents
the lexicon of one agent.}}
\end{figure}

@@

\begin{figure}[b]
\centerline{\includegraphics[width=.9\textwidth]{chap12/figs/strategies-one-winner.pdf}}
\caption{{Communicative success stays constant. The Brightness Colour Strategy becomes 
dominant very early due to random fluctuations. It stays dominant but barely. 
The strategy coherence (i.e. whether agents agree on the 
default strategy) increases gradually. The Full Colour Strategy remains still in use for cases where
it is needed. \label{fig:strategies-one}}}
\end{figure}

@@

\begin{figure}[b]
\centerline{\includegraphics[width=.9\textwidth]{chap12/figs/strategies-competition.pdf}}
\caption{{Initially the brightness strategy is dominant, but then the full-colour 
strategy gradually takes over and becomes dominant around 400 games/agents. Strategy coherence within the population
progressively increases.}
\label{fig:strategies-dynamics}}
\end{figure}

@@

The previous section showed how there can be competition or cooperation between strategies, fought out through the 
performance of the language systems against the niche factors in which \is{strategy generation}
the language strategies operate. But a Darwinian evolutionary framework requires that there is not only 
selection and a self-enforcing causal feedback loop 
but also that new elements get generated, in this case new strategies, which can then be subjected to 
a selectionist dynamics. How can we explain that new 
strategies arise? This is the most challenging difficult question of evolutionary linguistics research and very much an 
open question. We basically have to move up to a meta-level where we have strategies for building 
strategies, further called meta-strategies. \is{meta-strategy} It is unclear whether they have to be innately given or 
whether they are themselves also the result of developmental processes and a cultural dynamics. 
There are two aspects to a strategy: a conceptual side and a linguistic side. For the conceptual 
side, solid results have been achieved using IRL, which was from the beginning developed to allow the generation of 
new conceptualisation strategies. So far a single meta-strategy has proven to be quite successful in this case. 
The linguistic side however is still to a large extent in a stage of fundamental research, and here there appear to 
be several different meta-strategies, used by different language communities. Each of these two aspects is now 
discussed.

@@

\begin{figure}
\centerline{\includegraphics[width=.9\linewidth]{chap11/figs/near-the-box2.pdf}}
\caption{Example of a conceptualisation for the phrase `near to the box'. {\itshape Top:} 
Leftmost: The situation as perceived by the robot. Middle: situation after a geometric transform where the box is 
made the central perspective. Right: Area around the box which reflects (in terms of darkness of the color) how 
near objects are to the box. {\itshape Bottom:} IRL constraint network for this conceptualisation.
\label{fig:spatial-strategies}}
\end{figure}

We will use the domain of spatial language again as a source of examples, based on experiments carried out by 
Michael Spranger. It uses the \textsc{qrio} humanoid robots 
and scenes like the one shown in \figref{fig:spatial}. An example of a constraint network from this domain 
is shown in \figref{fig:spatial-strategies}. It is the network for the phrase `near the box'. `Near' is a proximal 
relation which categorises objects in terms of how near they are to a landmark (in this case the box). In a list form the 
network looks like this: 
\begin{verbatim}
1. (get-context ?ctx-1) 
2. (construct-region-proximal 
    ?region ?ctx-1 ?landmark ?cat) 
3. (bind proximal-category ?cat NEAR) 
4. (apply-selection ?landmark ?landmarks ?selector-2)
5. (apply-class ?landmarks ?ctx ?landmark-class) 
6. (bind selector ?selector-2 UNIQUE) 
7. (get-context ?ctx) 
8. (bind object-class ?landmark-class BOX)
\end{verbatim}
The order of these operations does not matter as values propagate and constraints become active until no further progress
can be made. Operations in 4, 5, 6, 7 and 8 provide the referent of `the box'. The following contributions 
can be made by each of the cognitive operations: The predicate BOX gets bound to ?landmark-class in operation 8. Object-class is 
the type of ?landmark-class. The context, i.e. all objects in the present situation, gets bound to ?ctx in operation 7. 
The selector UNIQUE gets bound to ?selector-2 in operation 6. 
Then the operation apply-class is applied to all the objects in ?ctx (the context) to find the possible landmarks which 
get bound to ?landmarks using the ?landmark-class (BOX). This set is assumed to be a singleton and so 
in operation 4 the unique element in the set of ?landmarks can be selected 
and bound to ?landmark. 
\enlargethispage{1\baselineskip}
The remaining three operations apply the proximal relation. In operation 3 ?cat is bound to the proximal-category NEAR. 
In operation 1, ?ctx-1 gets bound to 
the context. And finally in operation 2, the region around the landmark, which satisfies the category ?cat (NEAR), can 
be computed and bound to ?region by ``construct-region-proximal''. The construction-region-proximal constraint has to perform first 
a geometric transformation on the scene, constructing a world model from the viewpoint of the landmark, i.e. the box.

@@

\figref{fig:all} shows an example experiment in which all the aspects discussed in this section are brought together. 
A population of 10 agents is developing new conceptualisation strategies for a niche that encourages preference 
for a particular strategy, namely a projective strategy with possible perspective reversal, as in `the box in front 
of the wall'. We see that communicative success steadily rises and that the number of 
spatial categories increases. The bottom graph zooms in on the conceptualisation strategies. We see that agents reach 
a consensus about a dominant strategy after a period in which new strategies 
are invented and tried out.

\subsection{Meta-strategies for generating new lexico\-gram\-matical strategies}
% {\bfshape Meta-strategies for generating new lexico-grammatical strategies}\\

Are there general meta-strategies that can be used to generate the kind of lexico\-gram\-mat\-ical strategies 
that can in turn generate all the language systems we observe 
in today's languages? This is a very profound question, related to the ongoing debate in linguistics on the existence of 
a universal language acquisition device. A lot of ink has been spilled but we need to do profound computational modeling 
of meta-strategies and thorough experiments to find out.

Generally speaking, a lexico-grammatical strategy has two components: {\itshape diagnostics} which monitor how 
language processing is going and {\itshape repairs} which add new 
constructions to the agent's inventory.

@@

\enlargethispage{1\baselineskip}
It is not really possible to capture the enormous creativity of human word creation in computational simulations 
because that would involve components for common sense inference and access to massive cultural knowledge. 
However we can do experiments to understand how new words may arise by analogy with existing words 
and how the meaning of words can develop further, for example by abandoning some of the aspects of 
the original meaning or gaining new meanings. Such experiments have 
been carried out in particular by \cite{Wellens:2008}. \is{flexible word meaning}

@@

First we need much more work along the paths of which we have seen many examples: 
\begin{enumerate}
\item The study of more language systems, based on firm empirical data. For example, the study of the verbal system of 
Italian or the determiner system of Dutch. 
\item A study of the language strategies that underly these systems. A language strategy should define how a particular 
type of language system operates, how it is learned, and how it is expanded and aligned. 
\item A study of the niche occupied by the language systems governed by a particular language strategy. 
\item Evolutionary experiments to show how new strategies may arise. 
\end{enumerate} 
Then there are also several deep open questions which have not been addressed extensively so far. Here are a few examples: 
\begin{enumerate}
\item What are implications of the models presented here for the neurobiology of language? Right now it remains largely 
a mystery how the human brain is processing language and current neural network models are entirely insufficient from 
a representational and computational point of view to handle the structures and processes that we have been using in 
the various experiments reported in this book. 
\item How to orchestrate long-term evolution, both within the individual and in a population? Some suggestions have been 
made based on the autotelic principle which allows agents to control the communicative challenges in 
relation to their level of skill \citep{Steels:2007}. 
But these suggestions need to be tested
more seriously in large-scale long term experiments. 
\item What is the interaction between cultural evolution and genetic evolution? I have argued in favour of culture-driven genetic
evolution but what kind of genetic change is necessary to support the processes that we have been found to 
be operating at the cultural level? Clearly, flexible recruitment and huge amounts of memory are key demands on the 
human brain, but there are undoubtly others. 
\item What is the interaction between social evolution, cultural evolution and biological evolution? The language 
games which we have studied all required honest communication. Agents do their best to play the game as well as they 
can and agents are fully cooperative. This is not the default condition in the natural world and it is not so easy 
to explain how human sociality arose, and how human language may have fostered it \citep{Dor:2014}. 
\end{enumerate}
These questions can only be properly addressed by bringing up and working out hypotheses and
doing detailed computational analysis and computational 
experiments. The field of evolutionary linguistics is still in its infancy and so far too few researchers
are contributing through the kind of computational simulations that were discussed in this book. This is not surprising given 
the technical complexity and the breadth of knowledge needed to work in this area. It takes Ph.D students on average 5
years of training and experimentation to carry a research thread to 
fruition, assuming that the student starts with considerable background in computational linguistics or Artificial Intelligence.
Let us hope that in the future the resources will be available so that younger generations can pursue this exciting path. 
Many discoveries are waiting to be made!
